---
title: "Section 4: Population Scale Analysis"
author: Dani Baur (A16648266)
format: pdf
---

**Q13:** Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378.txt")
head(expr)
```
Sample size # for each genotype:
```{r}
table(expr$geno)
```
Sample size % for each genotype:
```{r}
table(expr$geno) / nrow(expr)
```
```{r}
df <- data.frame(expr)
zlot <- boxplot(df$exp ~ df$geno, notch=TRUE)
```
```{r}
zlot$stats
```
Median expression level for each genotype:
```{r}
zlot$stats[3,]
```
**Q14:** Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
z <- ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
z
```

Based on this boxplot, there is a notable difference between the expression levels of A|A vs G|G, meaning that the SNP affects the expression of ORMDL3. This means that when the genotype is homozygous A, there are higher levels of expression of ORMDL3 while homozygous G has lower levels of expression of ORMDL3. Also, heterozgyous A|G has an expression level in between these values.