---
title: "Class 8: Breast Cancer Mini Project"
author: Dani Baur (A16648266)
format: pdf
---

## About

In today's lab we will work with fine needle aspiration (FNA) of breast mass data from the University of Wisconsin.

## 1. Exploratory Data Analysis

```{r}
# data input
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```
```{r}
dim(wisc.df)
```
**Q:** How many patients/individuals/samples are in the dataset?
```{r}
nrow(wisc.df)
```

**Q:** How many of the observations have a malignant diagnosis?
```{r}
table(wisc.df$diagnosis)
```

**Q:** How many variables/features in the data are suffixed with _mean?
```{r}
ncol(wisc.df)
colnames(wisc.df)
inds <- grep("_mean", colnames(wisc.df))
length(inds)
```

Before analysis I want to take out the expert diagnoses column (aka the answer) from our dataset.
```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
head(diagnosis)
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

## 2. Principal Component Analysis

We can try a kmeans() clustering first
```{r}
km <- kmeans(wisc.data, centers=2)
km$cluster
```
```{r}
table(km$cluster)
```

Cross-table:
```{r}
table(km$cluster, diagnosis)
```

Let's try `hclust()` the key input required for `hclust()` is a distance matrix as produced by the `dist()` function.
```{r}
hc <- hclust(dist(wisc.data))
```

I can make a tree like figure:
```{r}
plot(hc)
```

Do we need to scale the data? 

We can look at the sd of each column (original value)
```{r}
round(apply(wisc.data, 2, sd))
```

Yes, we need to scale. We will run `prcomp()` with `scale=TRUE`.
```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Generate our main PCA plot (score plot, PC1 vs. PC2 plot)...
```{r}
library(ggplot2)

res <- as.data.frame(wisc.pr$x)

ggplot(res) +
  aes (x=PC1, y=PC2, col=diagnosis) +
  geom_point()
```

## 5. Clusting on PCA Results

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.

```{r}
d <- dist(wisc.pr$x[,1:3])
hc <- hclust(d, method="ward.D2")
plot(hc)
```
To get my clustering result/membership vector I need to "cut" the tree with the `cutree()` function.

```{r}
grps <- cutree(hc, h=80)
```

**Q.** How many patients are in each cluster group?

```{r}
table(grps)
```

```{r}
plot(res$PC1, res$PC2, col=grps)
```

## 7. Prediction

We can use our PCA result (model) to do predictions, that is take new unseen data and project it onto our new PC variables.

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(res$PC1, res$PC2, col=grps)
points(npc[,1],npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
**Q18.** Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 2 since the prediction shows similarity to the malignant results.

# Summary

Principal Component Analysis (PCA) is a super useful method for analyzing large datasets. It works by finding new variables (PCs) that capture the most variance from the original variables in your dataset.
